

# Methods {-}

## Engagement {-}

From May 2021 to the time of reporting SERNbc and McLeod Lake staff have been actively engaging with the following groups to discuss the project, solicit input, prioritize sites and plan fish passage remediation.

  * McLeod Lake Indian Band members of council
  * BCTS Engineering
  * CN Rail
  * Canadian Forest Products (Canfor)
  * Sinclar Forest Projects Ltd. (Sinclar)
  * Northern Engineering - Forests, Lands, Natural Resource Operations and Rural Development (FLNR)
  * BC Ministry of Transportation and Infrastructure
  * Fish Passage Technical Working Group
  * Planning foresters and biologists Ministry of Forests, Lands, Natural Resource Operations and Rural Development (restructured into Ministry of Forests and Ministry of Land, Water and Resource Stewardship)
  *  Fisheries experts
   


### Environmental Stewardship Initiative {-}

The [Environmental Stewardship Initiative](https://www2.gov.bc.ca/gov/content/environment/natural-resource-stewardship/consulting-with-first-nations/collaborative-stewardship-bc/environmental-stewardship-initiative)(ESI) is a collaborative partnership between the Province and First Nations with projects designed to focus on ecosystem assessment, restoration, enhancement, research and education . To date, four regional environmental stewardship projects have been developed. A governance working group with representatives from all four stewardship forums provides governance oversight and guidance for the ESI. The Parsnip River watershed group is within the Regional Strategic Environmental Assessment Forum (RSEA) area.

<br>

Working together with McLeod Lake Indian Band we are leveraging RSEA initiatives by overlaying Parsnip River watershed group fish passage planning data with the [Recreational Opportunity Spectrum Inventory data](https://catalogue.data.gov.bc.ca/dataset/recreational-opportunity-spectrum-inventory) to facilitate discussions with partners and stakeholders about how long term landscape stewardship goals of McLeod Lake Indian Band relate to stream crossing infrastructure upgrades/removals within their traditional territory.  The inventory data summarizes recreation opportunities as combinations of settings and probable human experience opportunities organized along a spectrum of classes. The spectrum is set out in terms of seven classes as follows: Primitive (P), Semi-primitive Non-motorized (SPNM), Semi-primitive Motorized (SPM), Roaded Natural (RN), Roaded Modified (RM), Rural (R), Urban (U).  The general idea is that areas towards the primitive end of the spectrum provide more opportunities for First Nations groups such as the McLeod Lake Indian Band to exercise their rights as indigenous people in pristine areas within their traditional territory [@unitednationsgeneralassembly2007UnitedNations].


## Habitat Modelling {-}

Through this initiative as well as other SERNbc led initiatives [@irvine2021BulkleyRiver; @irvine2022BulkleyRiver], as well as through direction from the Provincial Fish Passage Remediation Program and connectivity restoration planning conducted by Canadian Wildlife Federation and others [@mazany-wright_etal2021BulkleyRiver; @irvine2022ElkRiver], `bcfishpass` has been designed to prioritize potential fish passage barriers for assessment or remediation. The software is under continual development and has been designed and constructed by @norris2021smnorrisbcfishpass using of sql and python based shell script libraries to generate a simple model of aquatic habitat connectivity.  The model identifies natural barriers (ex. steep gradients for extended distances) and hydroelectric dams to classifying the accessibility upstream by fish [@norris2021smnorrisbcfishpass]. On potentially accessible streams, scripts identify known barriers (ex. waterfalls >5m high) and additional anthropogenic features which are primarily road/railway stream crossings (i.e. culverts) that are potentially barriers. To prioritize these features for assessment or remediation, scripts report on how much modelled potentially accessible aquatic habitat the barriers may obstruct. The model can be refined with known fish observations upstream of identified barriers and for each crossing location, the area of lake and wetland habitat upstream, species documented upstream/downstream, an estimate of watershed area (on 2nd order and higher streams), mean annual precipitation weighted to upstream watershed area and channel width can be collated using `bcfishpass`, `fwapg` and `bcfishobs`.  This, information, can be used to provide an indication of the potential quantity and quality of habitat potentially gained should fish passage be restored by comparing to user defined thresholds for the aforementioned parameters. 
 
 <br>
 
 `bcfishpass` calculates the average gradient of BC Freshwater Atlas stream network lines at minimum 100m long intervals starting from the downstream end of the streamline segment and working upstream.  The network lines are broken into max gradient categories with new segments created if and when the average slope of the stream line segment exceeds user provided thresholds. For this project, the user provided gradient thresholds used to delineate "potentially accessible habitat" were based on estimated max gradients that rainbow trout (20%) and bull trout (25%) are likely to be capable of ascending.  

<br>

Gradient, channel size and stream discharge are key determinants of channel morphology and subsequently fish distribution. High value  rearing, overwintering  and spawning habitat preferred by numerous species/life stages of fish are often located within channel types that have relatively low gradients and large channel widths (also quantified by the amount of flow in the stream).  Following delineation of "potentially accessible habitat", the average gradient of each stream segment within habitat classified as below the 20% and 25% thresholds was calculated and summed within species and life stage specific gradient categories.  Average gradient of stream line segments can be calculated from elevations contained in the provincial freshwater atlas streamline dataset. To obtain estimates of channel width upstream of crossing locations, where available, `bcfishpass` was utilized to pull average channel gradients from Fisheries Information Summary System (FISS) site assessment data [@moeStreamInventorySample] or PSCIS assessment data [@moe2021PSCISAssessments] and associate with stream segment lines. When both FISS and PSCIS values were associated with a particular stream segment, FISS channel width was used.  When multiple FISS sites were associated with a particular stream segment a mean of the average channel widths was taken.  To model channel width for 2nd order and above stream segments without associated FISS or PSCIS sites, first `fwapg` was used to estimate the drainage area upstream of the segment.  Then, rasters from ClimateBC [@wang_etal2012ClimateWNAHighResolution] were sampled for upstream watershed areas associated with each stream segment and a mean annual precipitation weighted by upstream watershed area was calculated. This data was used to model channel width using Bayesian statistical methods analyzing the relationship between input variables (n = 24849).  The details of this analysis and resulting formula used to estimate channel width on stream segments is included as a technical appendix `r if(gitbook_on){knitr::asis_output("[here](https://www.poissonconsulting.ca/f/859859031)")} else knitr::asis_output("at https://www.poissonconsulting.ca/f/859859031")`. 

<br>

`bcfishpass` and associated tools have been designed to be flexible in analysis, accepting user defined gradient, channel width and stream discharge categories [@moeStreamInventorySample].  Although currently in draft form, and subject to development revisions, gradient and channel width thresholds for habitat with the highest intrinsic value for a number of fish species in the Parsnip River watershed group have been specified and applied to model habitat upstream of stream crossing locations with the highest potential intrinsic value (Table \@ref(tab:tab-fish-spawning-rearing)).  Thresholds were derived based on a literature review with references provided in Table \@ref(tab:tab-fish-spawning-rearing-references). Output parameters for modelling are presented in Table \@ref(tab:tab-bcfp-def).



`r if(gitbook_on){knitr::asis_output("<br>")} else knitr::asis_output("<br><br><br>")`


```{r tab-fish-spawning-rearing, eval=F}
#`r if(identical(gitbook_on, FALSE)){knitr::asis_output("<br><br><br>")}`
bcfishpass_spawn_rear_model %>% 
  mutate(Species = fishbc::fbc_common_name(species_code), 
         spawn_gradient_max = round(spawn_gradient_max * 100 ,1),
         rear_gradient_max = round(rear_gradient_max * 100 ,1)) %>%
  select(Species, 
         `Spawning Gradient  Max (%)`= spawn_gradient_max,
         `Spawning Width Min (m)` = spawn_channel_width_min,
         # `Spawning Width Max (m)` = spawn_channel_width_max,
         # `Spawning MAD Min (m3/s)` = spawn_mad_min,
         # `Spawning MAD Max (m3/s)` = spawn_mad_max,
         `Rearing Gradient Max (%)` = rear_gradient_max) %>% 
         # `Rearing MAD Min (m3/s)` = rear_mad_min,
         # `Rearing MAD Max (m3/s)` = rear_mad_max,
         # `Rearing Wetland Multiplier` = rear_wetland_multiplier,
         # `Rearing Lake Multiplier` = rear_lake_multiplier) %>% 
  t() %>% 
  as_tibble(rownames = "row_names") %>% 
  janitor::row_to_names(row_number = 1) %>% 
  rename(Variable = Species) %>% 
  fpr::fpr_kable(caption_text = 'Stream gradient and channel width thresholds used to model potentially highest value fish habitat.', scroll = F)

```

<br>

```{r tab-fish-spawning-rearing-references, eval = F}
# bcfishpass_spawn_rear_model_references <- readr::read_csv(file = 'data/width_modelling/model_spawning_rearing_habitat.csv')
bcfishpass_spawn_rear_model_references <- readr::read_csv(file = 'data/inputs_raw/model_spawning_rearing_habitat.csv') 
  # select(species_code, contains('ref'), -contains(c('multiplier','mad')))
  
bcfishpass_spawn_rear_model_references %>% 
  mutate(Species = fishbc::fbc_common_name(species_code)) %>% 
  select(Species, 
         `Spawning Gradient  Max (%)`= spawn_gradient_max,
         `Spawning Width Min (m)` = spawn_channel_width_min,
         # `Spawning Width Max (m)` = spawn_channel_width_max_ref,
         # `Spawning MAD Min (m3/s)` = spawn_mad_min,
         # `Spawning MAD Max (m3/s)` = spawn_mad_max,
         `Rearing Gradient Max (%)` = rear_gradient_max) %>% 
         # `Rearing Wetland Multiplier` = rear_wetland_multiplier,
         # `Rearing Lake Multiplier` = rear_lake_multiplier) %>% 
         # `Rearing MAD Min (m3/s)` = rear_mad_min,
         # `Rearing MAD Max (m3/s)` = rear_mad_max) %>% 
  t() %>% 
  as_tibble(rownames = "row_names") %>% 
  janitor::row_to_names(row_number = 1) %>% 
  rename(Variable = Species) %>% 
  fpr::fpr_kable(caption_text = 'References for stream gradient and channel width thresholds used to model potentially highest value fish habitat. Preliminary and subject to revisions.', scroll = F)
```


<br>

```{r tab-bcfp-def, eval = F}
xref_bcfishpass_names %>% 
  filter(id_side == 1) %>% 
  arrange(id_join) %>%  
  select(Attribute = report, Definition = column_comment) %>% 
  fpr::fpr_kable(caption_text = 'bcfishpass outputs and associated definitions',
                 footnote_text = 'Steelhead model uses a gradient threshold of maximum 20% to determine if access if likely possible',
                 scroll = gitbook_on)

```


## Online Mapping {-}

An online mapping portal was


